---
title: "Aluminum Data"
author: "Van Myers"
date: 2020-02-12
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|--------------|----------------------------|-------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plates of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle

<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

### **q1** Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r regex-learning}
# # start with vignette
# who %>% 
#   pivot_longer(
#     cols = new_sp_m014:newrel_f65,
#     names_to = c("diagnosis", "gender", "age"), 
#     names_pattern = "new_?(.*)_(.)(.*)",
#     values_to = "count"
#   )
# 
# # adapt to this data
# df_stang %>% 
#   pivot_longer(
#     cols = c("E_00","E_45","mu_00","mu_45","E_90","mu_90"),
#     names_to = c(".value","angle"), 
#     names_pattern = "(.*)_(..)"
#   )
```

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
df_stang %>% 
  pivot_longer(
    cols = c("E_00","E_45","mu_00","mu_45","E_90","mu_90"),
    names_to = c(".value","angle"), 
    names_pattern = "(.*)_(..)",
    names_transform = c(E = NULL, nu = NULL, angle = as.integer)
  ) %>%
  filter(E>0)

## TODO: Is there a more elegant solution to renaming columns here?
names(df_stang_long)[names(df_stang_long) == 'mu'] <- 'nu'

df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA

<!-- -------------------------------------------------- -->

## Initial checks

<!-- ------------------------- -->

### **q2** Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
##
df_stang_long

df_stang_long %>%
  distinct(alloy)

df_stang_long %>%
  group_by(thick) %>%
  summarise(E = mean(E), nu = mean(nu))
```

**Observations**:

-   Is there "one true value" for the material properties of Aluminum?
    -   It appears there is no "one true value" for either property, but we can not draw this conclusion confidently. This dataset contains so few observations that any small differences are likely beyond statistical significance. What is clear is that the thickest sheet displays different material properties than the thinner sheets in these tests.
    -   Knowing what I do about aluminum (we don't define these properties based on thickness) I personally assume there may be a limit to the measurement setup that introduced an error.
-   How many aluminum alloys are in this dataset? How do you know?
    -   There is only one distinct alloy, 2024 (24ST). I manually inspected the data and confirmed there was one value for alloy with code.
-   What angles were tested?
    -   0, 45, and 90 deg
-   What thicknesses were tested?
    -   .022, .032, .064, and .081 inches
-   Are any of the observations outliers assuming both properties are intensive?

## Visualize

<!-- ------------------------- -->

### **q3** Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
## TASK: Investigate your question from q1 here
df_stang_long %>%
  ggplot(aes(x = E)) +
    geom_boxplot()

df_stang_long %>%
  ggplot(aes(x = E)) +
    geom_density()

df_stang_long %>%
  ggplot(aes(x = nu)) +
    geom_boxplot()

df_stang_long %>%
  ggplot(aes(x = nu)) +
    geom_density()
```

**Observations**:

-   There are no statistical outliers of E or nu in the dataset. However, the distribution of `E` is noticeably asymmetric. This may be explained by the small number of observations.

### **q4** Consider the following statement:

"A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

-   Does this graph support or contradict the claim above?
    -   This graph provides some support that `E` and `nu` are intensive properties as all observations for material thicknesses other than `.081` are clustered amongst each other. However, the observations for `.081` are separate which may suggest the properties are non-intensive. This still does not display a dependence on thickness and we have no statistical measure of significance to suggest a conclusive answer.
    -   From this graph I would lean towards saying both properties are intensive because there is no trend of dependence on thickness (but I already know they are).

# References

<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[<https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf>]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[<https://en.wikipedia.org/wiki/List_of_materials_properties>]
