---
title: "Gapminder"
author: "Van Myers"
date: 2020-2-13
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|--------------------------|----------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)
library(ggrepel)
library(gganimate)
library(av)
library(gifski)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

-   country
-   continent
-   year
-   lifeExp
-   pop
-   gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
df_year <- 
  gapminder %>%
  pull(year)

year_max <- max(df_year)
year_min <- min(df_year)
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = gdpPercap, color = continent)) +
    geom_density() +
  scale_x_log10()

gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = gdpPercap, color = continent)) +
    geom_boxplot() +
  scale_x_log10()

```

**Observations**:

-   Each continent has countries that vary greatly in GDP except Oceania.
    -   In this dataset Oceania is only Australia and New Zealand
-   Europe and the Americas are substantially wealthier than Africa and Asia in 1952

**Difficulties & Approaches**:

-   `geom_density` is not a particularly useful visualization because the outliers extend the axis beyond the bulk of the distribution.
    -   I initially filtered out the highest gdp country to address this
    -   Using a log scale on the `x` axis produces a better visualization, but the `geom_density` is still poorly suited to this task.
-   A boxplot is an overall better visualization. It handles the one extreme value very well but still benefits from the same filter to expand the useful real estate for the rest of the visualization.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

gapminder %>%
  group_by(continent) %>%
  filter(year == year_min) %>%
  mutate(outlier = find_outlier(gdpPercap)) %>%
  filter(outlier)
  
```

**Observations**:

-   Identify the outlier countries from q2
    -   Angola, Bahrain, Canada, Gabon, Kuwait, South Africa, Switzerland, United States, and Venezuela

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  group_by(continent) %>%
  filter(year == year_min) %>%
  mutate(outlier = find_outlier(gdpPercap)) %>%
  ggplot(aes(x = gdpPercap, y = continent, color = continent)) +
    geom_boxplot() + 
    geom_label_repel(
      data = . %>% filter(outlier),
      aes(label = country),
    ) +
    geom_label_repel(
      data = . %>% filter(country == "Kuwait"),
      aes(label = gdpPercap),
      nudge_y = -.35
    ) +
    coord_cartesian(xlim = c(0, 15000)) +
    theme(legend.position = "none")


gapminder %>%
  group_by(continent) %>%
  filter(year == year_max) %>%
  mutate(outlier = find_outlier(gdpPercap)) %>%
  ggplot(aes(x = gdpPercap, y = continent, color = continent)) +
    geom_boxplot() + 
    geom_label_repel(
      data = . %>% filter(outlier),
      aes(label = country),
      max.overlaps = 20
    ) +
    theme(legend.position = "none")
```

**Observations**:

-   Everyone is way richer
-   Africa is still really poor
-   Asia experienced the moist growth
-   The bulk of Europe joined the rest of the richest countries, which are less distant from poor countries than they were in the earliest year
-   Many outliers dissapeared. I'm not sure if this is a reduction in inequality or the US couping anyone that got too far out of place. Also some of the wealth may just have been unstable.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
gapminder %>%
  filter(country != "Kuwait") %>%
  filter(year == 2007) %>%
  ggplot(aes(x=lifeExp, y = gdpPercap, color = continent, label = country)) +
    geom_point() +
    geom_label_repel(show.legend = FALSE, max.overlaps = 10)

gapminder %>%
  filter(year == 2007) %>%
  arrange(lifeExp)
```

-   Obviously Kuwait is a huge outlier in terms of `gdpPercap`. There are 6 years where Kuwait has the highest `gdpPercap`, sometimes by a mile. I'll remove Kuwait to have a better view of everything else.
-   There are a handful of countries at the top of `gdpPercap`. Ireland comes as a small suprise to me, and there are a handful of Asian countries scattered throughout the top of GDP and life expectancy with the European colonial powers.
-   There are also a few African countries that stand out for having relatively high amounts of commercial activity relative to their population's life expectancy. This could be from especially high inequality (rich business people do lots of trade, but money doesn't reach the populace) or governmental corruption, or it could be *South Africa.*
-   There is a point around 70 years `lifeExp` below which countries are generally very poor. Past this point, we see a huge range of `gdpPercap`. The lower region is almost entirely African with a handful of Asian countries.
-   Afghanistan is the only Asian country well below 58 years `lifeExp` which makes sense with the United States doing United States things and all.

```{r q5-task2}
## TASK: Your second graph

# Calculation
gapminder_plus_stability <-
  gapminder %>%
  group_by(country) %>%
  mutate(
    lifeExp_stability = cor(year, lifeExp, method = "spearman"),
    gdpPercap_stability = cor(year, gdpPercap, method = "spearman"),
    rho_lifeExp_gdpPercap = cor(lifeExp, gdpPercap, method = "spearman"),
    total_stability = lifeExp_stability + gdpPercap_stability
    )

stability <-
  gapminder_plus_stability %>%
  filter(continent != "Oceania") %>%
  filter(year == 2007)

# Plotting

stability %>%
  ggplot(aes(x = continent, y = lifeExp_stability, color = continent)) +
    geom_boxplot(show.legend = FALSE)

stability %>%
  group_by(continent) %>%
  mutate(outlier = find_outlier(lifeExp_stability)) %>%
  ggplot(aes(y = lifeExp_stability, color = continent)) +
    geom_boxplot() +
    geom_label_repel(
      data = . %>% filter(outlier),
      aes(label = country, x = 0),
      max.overlaps = 12,
    ) +
  facet_wrap(~ continent) +
  theme(
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "none"
  )

# stability %>%
#   select(c(country, continent, lifeExp_stability, gdpPercap_stability)) %>%
#   arrange(lifeExp_stability)

stability %>%
  ggplot(aes(x = continent, y = gdpPercap_stability, color = continent)) +
    geom_boxplot(show.legend = FALSE)


stability %>%
  group_by(continent) %>%
  mutate(outlier = find_outlier(gdpPercap_stability)) %>%
  ggplot(aes(y = gdpPercap_stability, color = continent)) +
    geom_boxplot() +
    geom_label_repel(
      data = . %>% filter(outlier),
      aes(label = country, x = 0),
      max.overlaps = 12,
    ) +
  facet_wrap(~ continent) +
  theme(
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "none"
  )
# stability %>%
#   select(c(country, continent, gdpPercap_stability, lifeExp_stability)) %>%
#   arrange(gdpPercap_stability)

# stability %>%
#   select(
#     c(
#       country,
#       continent,
#       total_stability,
#       gdpPercap_stability,
#       lifeExp_stability
#       )
#     ) %>%
#   arrange(total_stability)
```

-   I've used the spearman correlation to describe an assumption I think is reasonable - I would expect a geopolitically stable country to experience an increasing `gdpPercap` and `lifeExp` as time passes. The spearman correlation tells us how true this is without looking at the rate of increase which varies depending on the state of development of a country. This would be worth investigating but seemed difficult with only 6 time points.
-   I am going to call this measure "stability". Stability is how consistently `lifeExp` and `gdpPercap` increase each year irrespective of how much of an increase occurs.
-   Africa is by far the most unstable continent in terms of both `lifeExp` and `gdpPercap`.
-   Outside of Africa, anything less than `1` is considered an outlier for `lifeExp_stability`.
-   `lifeExp` is far more stable than `gdpPercap`. This makes sense as economic turmoil is more common than the type of violence or poverty that would decrease life expectancy.
-   Because `lifeExp` is broadly more stable with less spread there are more outliers.
-   All outliers in Europe are Eastern European countries. This brings in to question the practice of continent grouping which is ultimately arbitrary as there are political and cultural regions that are likely more relevant.
-   Funny enough, every single country in Asia and the Americas that is an outlier in terms of `gdpPercap_stability` was couped or invaded by the US!
-   Eastern Europe is also the US's fault but there were a number of separate events (largely the rise of capitalism and balkanization)
-   Few countries in Africa and the Americas experienced consistent year over year growth in `gdpPercap`. Feels like another imperialism/colonialism moment.

```{r q5-task3}
## TASK: Your third graph

stability %>%
  ggplot(aes(
    x = gdpPercap_stability,
    y = lifeExp_stability,
    color = continent
    )) +
  geom_point()

stability %>%
  filter(gdpPercap_stability == 1 & lifeExp_stability < 1)
```

-   While making some of these plots I noticed a limit to my methodology. The spearman correlation doesn't weight a single precipitous drop as highly as a couple years of gradual decline. This means we will notice certain types of instability but might not highlight major turmoil if the country recovers quickly.
-   This last plot really should be in a separate section or the others should be removed, by I like them all.
-   A lot of countries have consistently increasing `lifeExp` (stability = 1) but less stable `gdpPercap`.
-   Other than the concentration at `lifeExp_stability = 1` the distribution of points appears to be fairly random.
-   Only one country has perfect `gdpPercap_stability` but poor (\<.9) `lifeExp_stability` . Three of the four countries where `gdpPercap_stability = 1` but `lifeExp_stability < 1` are Nordic countries which are known for their extreme wealth (from exploiting the third world) and strong health. This may represent them stagnating because they are waiting on new technology to increase lifespan and almost everything we can achieve with current technology is already in place there. Peurto Rico also shows up which I can't quite explain but they also enjoy a high life expectancy despite being less wealthy. Also natural disaster could have come into play.

```{r q5-additional-exploration}

stability %>%
  group_by(continent) %>%
  mutate(outlier = find_outlier(total_stability)) %>%
  ggplot(aes(x = total_stability, color = continent)) +
    geom_density()

stability %>%
  group_by(continent) %>%
  mutate(outlier = find_outlier(total_stability)) %>%
  ggplot(aes(x = total_stability, color = continent)) +
    geom_density() +
    geom_point(
      aes(x = total_stability, y = 0),
      size = 1
    ) +
    geom_point(
      data = . %>% filter(outlier),
      aes(x = total_stability, y = 0),
      size = 3
    ) # +
    # geom_label_repel(
    #     data = . %>% filter(outlier),
    #     aes(label = country, x = total_stability, y = 0),
    # )

stability %>%
  group_by(continent) %>%
  mutate(outlier = find_outlier(gdpPercap_stability)) %>%
  ggplot(aes(x = gdpPercap_stability, color = continent)) +
    geom_density()

stability %>%
  group_by(continent) %>%
  mutate(outlier = find_outlier(gdpPercap_stability)) %>%
  ggplot(aes(x = gdpPercap_stability, color = continent)) +
    geom_density() +
    geom_point(
      aes(x = gdpPercap_stability, y = 0),
      size = 1
    ) +
    geom_point(
      data = . %>% filter(outlier),
      aes(x = gdpPercap_stability, y = 0),
      size = 3
    ) # +
    # geom_label_repel(
    #     data = . %>% filter(outlier),
    #     aes(label = country, x = gdpPercap_stability, y = 0),
    # )

# gapminder_plus_stability <-
#   gapminder %>%
#   filter(continent != "Oceana") %>%
#   group_by(country) %>%
#   mutate(
#     lifeExp_stability = cor(year, lifeExp, method = "pearson"),
#     gdpPercap_stability = cor(year, gdpPercap, method = "pearson"),
#     rho_lifeExp_gdpPercap = cor(lifeExp, gdpPercap, method = "pearson")
#     )


gapminder_plus_stability %>%
  ggplot(aes(x = year, y = lifeExp, group = country, color = continent)) +
    geom_line(
      data = . %>% filter(lifeExp_stability >= .8),
      size = .3
    ) +
  geom_line(
      data = . %>% filter(lifeExp_stability <= .8),
      size = 1.5,
    )

gapminder %>%
  ggplot(aes(x=lifeExp, y = gdpPercap, color = continent, label = country)) +
    geom_point() +
    geom_label_repel(
      data = . %>% filter(gdpPercap > 58000),
      show.legend = FALSE
    )

no_kuwait <-
  gapminder %>%
  filter(country != "Kuwait") %>%
  ggplot(aes(x=lifeExp, y = gdpPercap, color = continent, label = country)) +
    geom_point() +
    theme_minimal(base_size = 11) +
    labs(
      x = "Life Expectancy",
      y = "GDP Per Capita",
    )
no_kuwait
no_kuwait + facet_wrap(~ year)

no_kuwait +
  theme_minimal(base_size = 20)
ggsave(
  filename = "test.png",
  path = "c04-gapminder-assignment_files",
  plot = no_kuwait,
  width = 16, height = 9, units = "in", dpi = 120
)


anim <-
  no_kuwait +
  transition_states(
    year,
    transition_length = 2,
    state_length = 0,
    wrap = FALSE
  ) +
  ease_aes('linear') +
  ggtitle('{closest_state}')

```

```{r render-animation}

animate(
  anim,
  fps = 30, duration = 10,
  width = 16, height = 9, units = "in", res = 120,
  renderer = gifski_renderer()
)
```
